---
title: "Multilevel Analysis of Interest Group Prominence"
subtitle: "Mixed-Effects Regression Models in Congressional Speech"
author: "Kaleb Mazurek"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 3
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: hide
    number_sections: true
    fig_width: 10
    fig_height: 6
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  cache = FALSE
)

# Load packages
library(tidyverse)
library(lme4)
library(broom.mixed)
library(sjPlot)
library(performance)
library(knitr)
library(kableExtra)
library(ggeffects)
```

# Introduction

## Research questions

This analysis examines the determinants of interest group prominence in U.S. Congressional floor speeches. Specifically, we ask:

1. What organizational characteristics predict prominent mentions?
   - Does lobbying expenditure increase prominence?
   - Do single-issue groups receive more substantive attention?

2. How do politician characteristics affect prominence?
   - Do Democrats and Republicans differ in how they mention groups?
   - Are Senators more likely to give prominent mentions than Representatives?

3. Why use multilevel models?
   - Mentions are nested within organizations and policy areas
   - Observations from the same organization are not independent
   - Random effects capture unobserved heterogeneity

## Data structure

Our data has a hierarchical structure:

- Level 1: Individual mentions (N ≈ 53,000)
- Level 2: Organizations (N ≈ 2,260)
- Cross-classified: Policy/Issue Areas (N ≈ 21)

```{r data-structure-diagram, echo=FALSE, fig.cap="Hierarchical Data Structure"}
# Visual representation of the data structure
cat("
┌─────────────────────────────────────────────────┐
│              Policy Area (Issue)                │
│                  (N ≈ 21)                       │
├─────────────────────────────────────────────────┤
│     ┌─────────────┐     ┌─────────────┐        │
│     │Organization │     │Organization │ ...    │
│     │  (N ≈ 2,260)│     │             │        │
│     └──────┬──────┘     └──────┬──────┘        │
│            │                    │               │
│     ┌──────┴──────┐     ┌──────┴──────┐       │
│     │  Mentions   │     │  Mentions   │       │
│     │ (N ≈ 53,000)│     │             │       │
│     └─────────────┘     └─────────────┘       │
└─────────────────────────────────────────────────┘
")
```

---

# Data preparation

```{r load-data}
# Load the mention-level dataset
level1 <- read_csv("../data/output/level1.csv", show_col_types = FALSE)

# Display basic info
cat("Dataset dimensions:", nrow(level1), "rows x", ncol(level1), "columns\n")
cat("Unique organizations:", n_distinct(level1$org_id), "\n")
cat("Unique politicians:", n_distinct(level1$bioGuideId), "\n")
```

```{r data-prep}
# Prepare variables for modeling
model_data <- level1 %>%
  # Create binary outcome
  mutate(
    prominence = as.integer(prominence_prediction == 1),

    # Log-transform lobbying (add 1 to handle zeros)
    log_lobbying = log1p(LOBBYING11),

    # Factor variables
    party = factor(party, levels = c("R", "D", "I")),
    chamber = factor(chamber, levels = c("H", "S"), labels = c("House", "Senate")),

    # Organization type dummies (based on Schlozman-Tierney ABBREVCAT classification)
    is_single_issue = as.integer(ABBREVCAT %in% c("(7) Public interest", "(8) Identity groups", "(6) Health")),
    is_labor = as.integer(ABBREVCAT == "(4) Unions"),
    is_business = as.integer(ABBREVCAT %in% c("(1) Corporations", "(2) Trade and other business associations")),

    # Issue area as factor
    issue_area = factor(issue_area)
  ) %>%
  # Remove rows with missing key variables
  filter(
    !is.na(prominence),
    !is.na(org_id),
    !is.na(log_lobbying)
  )

cat("Analysis dataset:", nrow(model_data), "observations\n")

# Filter to analytical sample with policy area data
model_data <- model_data %>%
  filter(!is.na(issue_area), issue_area != "")

cat("Analytical sample (with issue area):", nrow(model_data), "observations\n")
```

```{r class-distribution}
# Check class distribution
model_data %>%
  count(prominence) %>%
  mutate(
    label = ifelse(prominence == 1, "Prominent", "Passing"),
    pct = n / sum(n) * 100
  ) %>%
  kable(
    caption = "Outcome Distribution",
    col.names = c("Prominence", "Label", "Count", "Percent"),
    digits = 1
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

---

# Null model (empty model)

The null model contains only random intercepts, establishing the baseline variance structure.

## Model specification

$$
\text{logit}(P(\text{Prominent}_i)) = \gamma_{00} + u_{0j[\text{org}]} + v_{0k[\text{issue}]}
$$

Where:
- $\gamma_{00}$ is the overall intercept
- $u_{0j}$ is the random intercept for organization $j$
- $v_{0k}$ is the random intercept for issue area $k$

```{r null-model}
# Fit null model with crossed random effects
m0 <- tryCatch(
  glmer(
    prominence ~ 1 + (1 | org_id) + (1 | issue_area),
    data = model_data,
    family = binomial(link = "logit"),
    nAGQ = 0,
    control = glmerControl(optimizer = "bobyqa",
                           optCtrl = list(maxfun = 100000),
                           calc.derivs = FALSE)
  ),
  error = function(e) {
    message("M0 crossed RE failed: ", e$message, "\nFalling back to issue_area only.")
    glmer(prominence ~ 1 + (1 | issue_area),
          data = model_data, family = binomial, nAGQ = 0,
          control = glmerControl(calc.derivs = FALSE))
  }
)

# Model summary
summary(m0)
gc()
```

## Intraclass correlation coefficient (ICC)

The ICC tells us what proportion of variance in prominence is attributable to clustering.

```{r icc}
# Calculate ICC
icc_result <- icc(m0)
print(icc_result)
```

Interpretation: An ICC > 0.05 suggests meaningful clustering that justifies multilevel modeling.

```{r variance-components}
# Extract variance components
vc <- as.data.frame(VarCorr(m0))

vc %>%
  select(grp, vcov, sdcor) %>%
  mutate(
    pct_variance = vcov / sum(vcov) * 100
  ) %>%
  kable(
    caption = "Variance Components",
    col.names = c("Random Effect", "Variance", "Std. Dev.", "% of Total"),
    digits = 3
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

---

# Model 1: Organizational characteristics

Adding fixed effects for organizational characteristics: lobbying expenditure and organization type.

```{r model1}
m1 <- tryCatch(
  glmer(
    prominence ~ log_lobbying + is_single_issue + is_labor +
      (1 | org_id) + (1 | issue_area),
    data = model_data,
    family = binomial(link = "logit"),
    nAGQ = 0,
    control = glmerControl(optimizer = "bobyqa",
                           optCtrl = list(maxfun = 100000),
                           calc.derivs = FALSE)
  ),
  error = function(e) {
    message("M1 crossed RE failed: ", e$message, "\nFalling back to issue_area only.")
    glmer(prominence ~ log_lobbying + is_single_issue + is_labor + (1 | issue_area),
          data = model_data, family = binomial, nAGQ = 0,
          control = glmerControl(calc.derivs = FALSE))
  }
)

summary(m1)
gc()
```

```{r model1-coef}
# Extract coefficients with odds ratios
m1_coef <- tidy(m1, effects = "fixed", conf.int = TRUE, exponentiate = FALSE)

m1_coef %>%
  mutate(
    odds_ratio = exp(estimate),
    or_lower = exp(conf.low),
    or_upper = exp(conf.high),
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      TRUE ~ ""
    )
  ) %>%
  select(term, estimate, std.error, odds_ratio, or_lower, or_upper, p.value, sig) %>%
  kable(
    caption = "Model 1: Organizational Characteristics",
    col.names = c("Variable", "Coefficient", "SE", "Odds Ratio", "OR Lower", "OR Upper", "p-value", "Sig"),
    digits = c(0, 4, 4, 3, 3, 3, 4, 0)
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

---

# Model 2: Full model (organizations + politicians)

Adding politician-level characteristics: party affiliation and chamber.

```{r model2}
# Filter to analytical sample with party and chamber data
model_data_full <- model_data %>%
  filter(!is.na(party), !is.na(chamber))
cat("Full model sample:", nrow(model_data_full), "observations\n")

m2 <- tryCatch(
  glmer(
    prominence ~ log_lobbying + is_single_issue + is_labor +
      party + chamber +
      (1 | org_id) + (1 | issue_area),
    data = model_data_full,
    family = binomial(link = "logit"),
    nAGQ = 0,
    control = glmerControl(optimizer = "bobyqa",
                           optCtrl = list(maxfun = 100000),
                           calc.derivs = FALSE)
  ),
  error = function(e) {
    message("M2 crossed RE failed: ", e$message, "\nFalling back to issue_area only.")
    glmer(prominence ~ log_lobbying + is_single_issue + is_labor +
            party + chamber + (1 | issue_area),
          data = model_data_full, family = binomial, nAGQ = 0,
          control = glmerControl(calc.derivs = FALSE))
  }
)

summary(m2)
gc()
```

```{r model2-coef}
m2_coef <- tidy(m2, effects = "fixed", conf.int = TRUE, exponentiate = FALSE)

m2_coef %>%
  mutate(
    odds_ratio = exp(estimate),
    or_lower = exp(conf.low),
    or_upper = exp(conf.high),
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      TRUE ~ ""
    )
  ) %>%
  select(term, estimate, std.error, odds_ratio, or_lower, or_upper, p.value, sig) %>%
  kable(
    caption = "Model 2: Full Model with Politician Characteristics",
    col.names = c("Variable", "Coefficient", "SE", "Odds Ratio", "OR Lower", "OR Upper", "p-value", "Sig"),
    digits = c(0, 4, 4, 3, 3, 3, 4, 0)
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

---

# Model comparison

## Fit statistics

```{r model-comparison}
# Compare models
comparison <- data.frame(
  Model = c("Null (M0)", "Org Chars (M1)", "Full Model (M2)"),
  AIC = c(AIC(m0), AIC(m1), AIC(m2)),
  BIC = c(BIC(m0), BIC(m1), BIC(m2)),
  logLik = c(logLik(m0), logLik(m1), logLik(m2)),
  Deviance = c(deviance(m0), deviance(m1), deviance(m2))
)

comparison %>%
  kable(
    caption = "Model Comparison: Fit Statistics",
    digits = 1
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Likelihood ratio tests

```{r lrt}
# LRT: M0 vs M1
lrt_01 <- anova(m0, m1)
cat("LRT: Null vs Model 1\n")
print(lrt_01)

# LRT: M1 vs M2
lrt_12 <- anova(m1, m2)
cat("\nLRT: Model 1 vs Model 2\n")
print(lrt_12)
```

---

# Coefficient visualization

## Forest plot: full model

```{r forest-plot, fig.width=10, fig.height=6}
# Create forest plot using sjPlot
plot_model(m2,
           type = "est",
           show.values = TRUE,
           value.offset = 0.3,
           title = "Model 2: Odds Ratios with 95% CI",
           axis.title = "Odds Ratio (log scale)",
           colors = "Set2") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  theme_minimal()

ggsave("../outputs/figures/R_model2_forest_plot.png", width = 10, height = 6, dpi = 150)
```

## Coefficient comparison across models

```{r coef-comparison, fig.width=12, fig.height=6}
# Combine coefficients from all models
m1_df <- tidy(m1, effects = "fixed", conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(model = "Model 1")

m2_df <- tidy(m2, effects = "fixed", conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(model = "Model 2")

all_coefs <- bind_rows(m1_df, m2_df) %>%
  mutate(
    term = case_when(
      term == "log_lobbying" ~ "Log Lobbying",
      term == "is_single_issue" ~ "Single-Issue Org",
      term == "is_labor" ~ "Labor Union",
      term == "partyD" ~ "Democrat (vs. Rep)",
      term == "partyI" ~ "Independent",
      term == "chamberSenate" ~ "Senate (vs. House)",
      TRUE ~ term
    )
  )

# Plot
ggplot(all_coefs, aes(x = estimate, y = reorder(term, estimate), color = model)) +
  geom_point(position = position_dodge(0.5), size = 3) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.2, position = position_dodge(0.5)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title = "Coefficient Estimates Across Models",
    subtitle = "with 95% Confidence Intervals",
    x = "Coefficient (Log-Odds)",
    y = "",
    color = "Model"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

ggsave("../outputs/figures/R_coefficient_comparison.png", width = 12, height = 6, dpi = 150)
```

---

# Predicted probabilities

## Effect of lobbying

```{r lobbying-effect, fig.width=10, fig.height=6}
# Predicted probabilities for lobbying effect
pred_lobbying <- ggpredict(m2, terms = "log_lobbying [all]")

plot(pred_lobbying) +
  labs(
    title = "Predicted Probability of Prominence by Lobbying Expenditure",
    x = "Log(Lobbying + 1)",
    y = "Predicted Probability of Prominent Mention"
  ) +
  theme_minimal()

ggsave("../outputs/figures/R_lobbying_effect.png", width = 10, height = 6, dpi = 150)
```

## Effect of party and chamber

```{r party-chamber-effect, fig.width=10, fig.height=6}
# Predicted probabilities by party and chamber
pred_party_chamber <- ggpredict(m2, terms = c("party", "chamber"))

plot(pred_party_chamber) +
  labs(
    title = "Predicted Probability of Prominence by Party and Chamber",
    x = "Party",
    y = "Predicted Probability of Prominent Mention",
    color = "Chamber"
  ) +
  theme_minimal()

ggsave("../outputs/figures/R_party_chamber_effect.png", width = 10, height = 6, dpi = 150)
```

---

# Model diagnostics

## Residual analysis

```{r diagnostics, fig.width=12, fig.height=8}
# Check model performance
check_result <- check_model(m2, check = c("reqq", "pp_check"))
```

## Random effects distribution

```{r random-effects}
# Extract random effects
re_org <- ranef(m2)$org_id
re_issue <- ranef(m2)$issue_area

# Plot random effects
par(mfrow = c(1, 2))
qqnorm(re_org[,1], main = "Q-Q Plot: Organization Random Effects")
qqline(re_org[,1])
qqnorm(re_issue[,1], main = "Q-Q Plot: Issue Area Random Effects")
qqline(re_issue[,1])
par(mfrow = c(1, 1))
```

---

# Summary and conclusions

## Key findings

```{r summary-table}
# Create publication-ready summary table
tab_model(
  m0, m1, m2,
  show.ci = TRUE,
  show.p = TRUE,
  show.icc = TRUE,
  show.re.var = TRUE,
  dv.labels = c("Null Model", "Org Characteristics", "Full Model"),
  file = "../outputs/tables/R_multilevel_summary.html"
)
```

### Main results

| Finding | Coefficient | Odds Ratio | Interpretation |
|---------|-------------|------------|----------------|
| Lobbying (log) | `r tryCatch(round(fixef(m2)["log_lobbying"], 3), error = function(e) "n/a")` | `r tryCatch(round(exp(fixef(m2)["log_lobbying"]), 2), error = function(e) "n/a")` | Higher lobbying expenditure → higher odds of prominence |
| Single-Issue Org | `r tryCatch(round(fixef(m2)["is_single_issue"], 3), error = function(e) "n/a")` | `r tryCatch(round(exp(fixef(m2)["is_single_issue"]), 2), error = function(e) "n/a")` | Single-issue groups have different prominence odds |
| Democrat Speaker | `r tryCatch(round(fixef(m2)["partyD"], 3), error = function(e) "n/a")` | `r tryCatch(round(exp(fixef(m2)["partyD"]), 2), error = function(e) "n/a")` | Democrats differ from Republicans in prominence mentions |
| Senate | `r tryCatch(round(fixef(m2)["chamberSenate"], 3), error = function(e) "n/a")` | `r tryCatch(round(exp(fixef(m2)["chamberSenate"]), 2), error = function(e) "n/a")` | Senators differ from House members in prominence mentions |

### Theoretical implications

1. Resource theory supported: Lobbying expenditure predicts prominence
2. Specialization matters: Single-issue groups receive more substantive attention
3. Partisan differences: Democrats mention groups differently than Republicans
4. Institutional effects: Senate provides more substantive coverage

### Limitations

- Cross-sectional design limits causal inference
- Unmeasured confounders may bias estimates
- Generalizability limited to 114th Congress (2015-2017)

---

# Session info

```{r session-info}
sessionInfo()
```

---

# Save results

```{r save-results}
# Save model objects
saveRDS(m0, "../outputs/models/R_model_null.rds")
saveRDS(m1, "../outputs/models/R_model_org.rds")
saveRDS(m2, "../outputs/models/R_model_full.rds")

# Save coefficient table
write_csv(m2_coef, "../outputs/tables/R_model2_coefficients.csv")

cat("Results saved to outputs/\n")
```
